---
title: "How to use the R2ucare package to assess the fit of capture-recapture to data?"
author: "Olivier Gimenez"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Session examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## What it does (and does not do)

Ths package contains R functions to perform goodness-of-fit tests for capture-recapture models. It also has various functions to manipulate capture-recapture data. Please email all suggestions for improvements, questions, comments and bugs to olivier.gimenez [AT] cefe.cnrs.fr.

For Cormack-Jolly-Seber models (single-state), we refer to Lebreton et al. (1992) and Pradel et al. (2005) for the theory. For Arnason-Schwarz models (multistate), have a look to Pradel et al. (2003). [Chapter 5 of the Gentle Introduction to MARK](http://www.phidot.org/software/mark/docs/book/pdf/chap5.pdf) also provides a good start for understanding goodness-of-fit test. 

**Warning**: to date, no goodness-of-fit test exists for models with individual covariates (unless you discretize them and use groups), individual time-varying covariates (unless you treat them as states) or temporal covariates; therefore, remove these covariates from your dataset before using it with R2ucare. For groups, just treat the group separately as in the Dipper example below. 

## To install the package

The latest stable version of the package can be downloaded from CRAN with the R command
``` 
install.packages("unmarked")
```

The repository on GitHub https://github.com/oliviergimenez/R2ucare hosts the development version of the package, to install it:
```
if(!require(devtools)) install.packages("devtools")
library("devtools")
install_github('oliviergimenez/R2ucare')
```

Despite what its name might suggest, **you do not need** to download and install U-CARE to run the R2ucare package. This package is basically a Matlab to R translation of U-CARE (Choquet et al. 2009). 

## Goodness-of-fit tests for the Cormack-Jolly-Seber model

First things first, load the `R2ucare` package:
```{r, message=FALSE, warning=FALSE}
library(R2ucare)
```

Let's begin by reading in the classical dipper using the `read_inp` function. Note that data with the Headed format can also be read using the function `read_headed`. The dataset is provided with the package (thanks to Gilbert Marzolin for sharing his data).

```{r, message=FALSE, warning=FALSE}
dipper = system.file("extdata", "ed.inp", package = "R2ucare")
dipper = read_inp(dipper,group.df=data.frame(sex=c('Male','Female')))
```

Get encounter histories, counts and groups:
```{r, message=FALSE, warning=FALSE}
dip.hist = dipper$encounter_histories
dip.freq = dipper$sample_size
dip.group = dipper$groups
```

Split the dataset in females/males:
```{r, message=FALSE, warning=FALSE}
mask = (dip.group == 'Female')
dip.fem.hist = dip.hist[mask,]
dip.fem.freq = dip.freq[mask]
mask = (dip.group == 'Male')
dip.mal.hist = dip.hist[mask,]
dip.mal.freq = dip.freq[mask]
```

Tadaaaaan, now you're ready to perform Test.3Sr, Test3.Sm, Test2.Ct and Test.Cl for females:
```{r, message=FALSE, warning=FALSE}
test3sr_females = test3sr(dip.fem.hist, dip.fem.freq)
test3sm_females = test3sm(dip.fem.hist, dip.fem.freq)
# we need the m-array to perform test2ct and test2cl
test2ct_females = test2ct(dip.fem.hist, dip.fem.freq)
test2cl_females = test2cl(dip.fem.hist, dip.fem.freq)
# display results:
test3sr_females
test3sm_females
test2ct_females
test2cl_females
```

Or perform all tests at once:
```{r, message=FALSE, warning=FALSE}
overall_CJS(dip.fem.hist, dip.fem.freq)
```

## Goodness-of-fit tests for the Arnason-Schwarz model

Read in the geese dataset. It is provided with the package (thanks to Jay Hesbeck for sharing his data). 

```{r, message=FALSE, warning=FALSE}
geese = system.file("extdata", "geese.inp", package = "R2ucare")
geese = read_inp(geese)
```

Get encounter histories and number of individuals with corresponding histories
```{r, message=FALSE, warning=FALSE}
geese.hist = geese$encounter_histories
geese.freq = geese$sample_size
```

And now, perform Test3.GSr, Test.3.GSm, Test3G.wbwa, TestM.ITEC and TestM.LTEC:
```{r, message=FALSE, warning=FALSE}
test3Gsr(geese.hist,geese.freq)
test3Gsm(geese.hist,geese.freq)
test3Gwbwa(geese.hist,geese.freq)
testMitec(geese.hist,geese.freq)
testMltec(geese.hist,geese.freq)
```

Or all tests at once:
```{r, message=FALSE, warning=FALSE}
overall_JMV(geese.hist,geese.freq)
```

## Various useful functions

Several U-CARE functions to manipulate and process capture-recapture data can be mimicked with R built-in functions. For example, recoding multisite/state encounter histories in single-site/state ones is easy:
```{r, message=FALSE, warning=FALSE}
# Assuming the geese dataset has been read in R (see above):
geese.hist[geese.hist>1] = 1
```
So is recoding states:
```{r, message=FALSE, warning=FALSE}
# Assuming the geese dataset has been read in R (see above):
geese.hist[geese.hist==3]=2 # all 3s become 2s
```
Also, reversing time is not that difficult:
```{r, message=FALSE, warning=FALSE,eval=FALSE}
# Assuming the female dipper dataset has been read in R (see above):
t(apply(dip.fem.hist,1,rev))
```
What about cleaning data, i.e. deleting empty histories, and histories with no individuals?
```{r, message=FALSE, warning=FALSE,eval=FALSE}
# Assuming the female dipper dataset has been read in R (see above):
mask = (apply(dip.fem.hist,1,sum)>0 & dip.fem.freq>0) # select non-empty histories, and histories with at least one individual
sum(!mask) # how many histories are to be dropped?
dip.fem.hist[mask,] # drop these histories from dataset
dip.fem.freq[mask] # from counts
```
Selecting or gathering occasions is as simple as that:
```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Assuming the female dipper dataset has been read in R (see above):
dip.fem.hist[,c(1,4,6)] # pick occasions 1, 4 and 6 (might be a good idea to clean the resulting dataset)
gather_146 = apply(dip.fem.hist[,c(1,4,6)],1,max) # gather occasions 1, 4 and 6 by taking the max
dip.fem.hist[,1] = gather_146 # replace occasion 1 by new occasion
dip.fem.hist = dip.fem.hist[,-c(4,6)] # drop occasions 4 and 6
```
Finally, suppressing the first encounter is achieved as follows:
```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Assuming the geese dataset has been read in R (see above):
for (i in 1:nrow(geese.hist)){
occasion_first_encounter = min(which(geese.hist[i,]!=0)) # look for occasion of first encounter
geese.hist[i,occasion_first_encounter] = 0 # replace the first non zero by a zero
}
# delete empty histories from the new dataset
mask = (apply(geese.hist,1,sum)>0) # select non-empty histories
sum(!mask) # how many histories are to be dropped?
geese.hist[mask,] # drop these histories from dataset
geese.freq[mask] # from counts
```

The R packages `plyr`, `dplyr` and `tidyr` might also be very useful to work with capture-recapture datasets.

Besides these simple manipulations, several useful U-CARE functions need a proper R equivalent. I have coded a few of them, see the list below and ?name-of-the-function for more details. 

* `marray` build the m-array for single-site/state capture-recapture data;
* `multimarray` build the m-array for multi-site/state capture-recapture data;
* `group_data` pool together individuals with the same encounter capture-recapture history.
* `ungroup_data` split encounter capture-recapture histories in individual ones.

## References 

* Choquet, R., Lebreton, J.-D., Gimenez, O., Reboulet, A.-M., and R. Pradel. (2009). [U-CARE: Utilities for performing goodness of fit tests and manipulating CApture-REcapture data](https://dl.dropboxusercontent.com/u/23160641/my-pubs/Choquetetal2009UCARE.pdf). Ecography. 32: 1071-1074.
* Lebreton, J.-D. et al. (1992). Modeling survival and testing biological hypotheses using marked animals: a unified approach with case studies. Ecol. Monogr. 62: 67-118.
* Pradel, R., Gimenez O. and J.-D. Lebreton (2005). [Principles and interest of GOF tests for multistate capture-recapture models](https://dl.dropboxusercontent.com/u/23160641/my-pubs/Pradeletal2005ABC.pdf). Animal Biodiversity and Conservation 28: 189–204.
* Pradel R., Wintrebert C.M.A. and Gimenez O. (2003). [A proposal for a goodness-of-fit test to the Arnason-Schwarz multisite capture-recapture model](https://dl.dropboxusercontent.com/u/23160641/my-pubs/Pradeletal2003Biometrics.pdf). Biometrics 59: 43-53.
